---
title: "STA141A Final Project"
author: "Daniel Cayapan"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/Users/danie/Downloads/')
```
Datasets For 2022-2023 Season
```{r}
nba <- read.csv("nba_salaries.csv")
print(head(nba))

nba1 <- read.csv("NBAWL.csv")
print(head(nba1))

nba2 <- read.csv("NBAData1.csv")
print(head(nba2))
```
Merge the three data sets
```{r}
library(dplyr)

merged_nba_full <- full_join(nba, nba2, by = "Team")
print(head(merged_nba_full)) # print out first couple rows

merged_nba_full <-full_join(merged_nba_full, nba1, by = "Team")
print(head(merged_nba_full))
```
Cleaning Data: IF NA salary due to player playing for multiple teams, just omit entire player. Also delete overlapping columns between datasets that has inaccurate or data that is not needed for this analysis.
```{r}
library(tidyr)

merged_nba_clean <- drop_na(merged_nba_full)

merged_nba_clean <- subset(merged_nba_clean, select = -c(Rank, RPI, W, L, PCT, SOS, PWR, Player.additional, PF.y, PA, Signed))
merged_nba_clean <- drop_na(merged_nba_clean)
print(head(merged_nba_clean))
```
```{r}
# load necessary libraries
library(dplyr)
library(ggplot2)
library(factoextra)

# Select the salary column
salary_data <- merged_nba_clean %>% select(Salary)

# Normalize the salary data
salary_data <- scale(salary_data)

# Elbow method to find the optimal number of clusters
fviz_nbclust(salary_data, kmeans, method = "wss") +
  geom_vline(xintercept = 3, linetype = 2) +
  labs(subtitle = "Elbow method")

# Set seed for reproducibility
set.seed(123)

# Apply K-means clustering
kmeans_result <- kmeans(salary_data, centers = 3, nstart = 25)

# Add cluster assignment to the original dataset
merged_nba_clean$cluster <- kmeans_result$cluster

# Plot the clusters
ggplot(merged_nba_clean, aes(x = Salary, y = Salary, color = factor(cluster))) +
  geom_point() +
  labs(title = "K-means Clustering of Players by Salary",
       x = "Salary",
       y = "Salary",
       color = "Cluster") +
  theme_minimal()

```
```{r}
# Count the number of players in each cluster
cluster_counts <- merged_nba_clean %>%
  group_by(cluster) %>%
  summarise(count = n())

print(cluster_counts)
```

```{r}
# Calculate average salary per cluster
avg_salary_per_cluster <- merged_nba_clean %>%
  group_by(cluster) %>%
  summarize(avg_salary = mean(Salary, na.rm = TRUE))

# Merge average salary per cluster with the original data
merged_nba_clean <- merge(merged_nba_clean, avg_salary_per_cluster, by = "cluster")

# Calculate the correlation between average salary in clusters and team win percentage
correlation1 <- cor(merged_nba_clean$avg_salary, merged_nba_clean$WP, use = "complete.obs")
print(correlation1)
```
The correlation is a weak positive relationship between the average salary of players in a given cluster to the team's win percentage in the 2022-2023 NBA season. This means there is a very slight tendency for teams to have a higher win percentage when paying their players more.

```{r}
# Summarize statistics by cluster
cluster_stats1 <- merged_nba_clean %>%
  group_by(cluster) %>%
  summarise(
    avg_salary = mean(Salary, na.rm = TRUE),
    avg_points_per_game = mean(PTS, na.rm = TRUE),
    avg_assists_per_game = mean(AST, na.rm = TRUE),
    avg_rebounds_per_game = mean(TRB, na.rm = TRUE),
    avg_steals_per_game = mean(STL, na.rm = TRUE),
    avg_blocks_per_game = mean(BLK, na.rm = TRUE)
    
  )

print(cluster_stats1)
```
```{r}
# Calculate correlation within each cluster
correlation_by_cluster <- merged_nba_clean %>%
  group_by(cluster) %>%
  summarize(
    correlation_salary_points = cor(Salary, PTS, use = "complete.obs"),
    correlation_salary_assists = cor(Salary, AST, use = "complete.obs"),
    correlation_salary_rebounds = cor(Salary, TRB, use = "complete.obs"),
    correlation_salary_steals = cor(Salary, STL, use = "complete.obs"),
    correlation_salary_blocks = cor(Salary, BLK, use = "complete.obs")
    
  )

# View correlation by cluster
print(correlation_by_cluster)
```
All correlations between clusters and player stats were relatively low, however what does stick out is cluster two with all stats having a slightly more moderate positive correlation between the cluster and the statistics. We note that cluster 2 has the most amount of players.

We are now going to cluster by Player statistics
```{r}
# Calculate correlation between each player statistic and win percentage
correlation_points <- cor(merged_nba_clean$PTS, merged_nba_clean$WP)
correlation_assists <- cor(merged_nba_clean$AST, merged_nba_clean$WP)
correlation_rebounds <- cor(merged_nba_clean$TRB, merged_nba_clean$WP)
correlation_steals <- cor(merged_nba_clean$STL, merged_nba_clean$WP)
correlation_blocks <- cor(merged_nba_clean$BLK, merged_nba_clean$WP)


# View correlation for each player statistic
print(paste("Correlation with Points:", correlation_points))
print(paste("Correlation with Assists:", correlation_assists))
print(paste("Correlation with Rebounds:", correlation_rebounds))
print(paste("Correlation with Steals:", correlation_steals))
print(paste("Correlation with Blocks:", correlation_blocks))
```
All stats provide a relatively weak relationship between individual players' stats and teams' win percentage.
```{r}
# Load necessary libraries
library(ggplot2)

# Create a data frame for plotting
correlation_data <- data.frame(
  Statistic = c("Points", "Assists", "Rebounds", "Steals", "Blocks"),
  Correlation = c(correlation_points, correlation_assists, correlation_rebounds, correlation_steals, correlation_blocks)
)

# Create a scatter plot
ggplot(correlation_data, aes(x = Statistic, y = Correlation)) +
  geom_point(size = 3, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Player Statistic", y = "Correlation with Win Percentage", title = "Correlation between Player Statistics and Win Percentage")

```






