---
title: "STA141A Final Project"
author: "Daniel Cayapan"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/Users/danie/Downloads/')
```
Datasets For 2022-2023 Season
```{r}
nba <- read.csv("nba_salaries.csv")
nba

nba1 <- read.csv("NBAWL.csv")
nba1
```
Merge the two data sets
```{r}
library(dplyr)

merged_nba_full <- full_join(nba, nba1, by = "Team")
print(head(merged_nba_full)) # print out first couple rows
```
Cleaning Data: IF NA salary due to player playing for multiple teams, just omit entire player.
```{r}
library(tidyr)

merged_nba_clean <- drop_na(merged_nba_full)
print(head(merged_nba_clean))
```
```{r}
# load necessary libraries
library(dplyr)
library(ggplot2)
library(factoextra)

# Select the salary column
salary_data <- merged_nba_clean %>% select(Salary)

# Normalize the salary data
salary_data <- scale(salary_data)

# Elbow method to find the optimal number of clusters
fviz_nbclust(salary_data, kmeans, method = "wss") +
  geom_vline(xintercept = 3, linetype = 2) +
  labs(subtitle = "Elbow method")

# Set seed for reproducibility
set.seed(123)

# Apply K-means clustering
kmeans_result <- kmeans(salary_data, centers = 3, nstart = 25)

# Add cluster assignment to the original dataset
merged_nba_clean$cluster <- kmeans_result$cluster

# Plot the clusters
ggplot(merged_nba_clean, aes(x = Salary, y = Salary, color = factor(cluster))) +
  geom_point() +
  labs(title = "K-means Clustering of Players by Salary",
       x = "Salary",
       y = "Salary",
       color = "Cluster") +
  theme_minimal()

```
```{r}
# Count the number of players in each cluster
cluster_counts <- merged_nba_clean %>%
  group_by(cluster) %>%
  summarise(count = n())

print(cluster_counts)
```

```{r}
# Calculate average salary per cluster
avg_salary_per_cluster <- merged_nba_clean %>%
  group_by(cluster) %>%
  summarize(avg_salary = mean(Salary, na.rm = TRUE))

# Merge average salary per cluster with the original data
merged_nba_clean <- merge(merged_nba_clean, avg_salary_per_cluster, by = "cluster")

# Calculate the correlation between average salary in clusters and team win percentage
correlation1 <- cor(merged_nba_clean$avg_salary, merged_nba_clean$WP, use = "complete.obs")
print(correlation1)
```
The correlation is a weak positive relationship between the average salary of players in a given cluster to the team's win percentage in the 2022-2023 NBA season. This means there is a very slight tendency for teams to have a higher win percentage when paying their players more.

```{r}

```

