---
title: "STA141A Final Project"
author: "Daniel Cayapan, Ian Dang, Yiming Cao, Wenzhuo Sun"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/Users/danie/Downloads/')
```
Datasets For 2022-2023 Season
```{r}
nba <- read.csv("nba_salaries.csv")
print(head(nba))

nba1 <- read.csv("NBAWL.csv")
print(head(nba1))

nba2 <- read.csv("NBAData1.csv")
print(head(nba2))
```
Merge the three data sets
```{r}
library(dplyr)

merged_nba_full <- full_join(nba, nba2, by = "Team")
print(head(merged_nba_full)) # print out first couple rows

merged_nba_full <-full_join(merged_nba_full, nba1, by = "Team")
print(head(merged_nba_full))
```
Cleaning Data: IF NA salary due to player playing for multiple teams, just omit entire player. Also delete overlapping columns between datasets that has inaccurate or data that is not needed for this analysis.
```{r}
library(tidyr)

merged_nba_clean <- drop_na(merged_nba_full)


merged_nba_clean <- subset(merged_nba_clean, select = -c(Rank, RPI, W, L, PCT, SOS, PWR, Player.additional, PF.y, PA, Signed))
merged_nba_clean <- drop_na(merged_nba_clean)

print(head(merged_nba_clean))
```
```{r}
# Fit a linear regression model
model <- lm(WP ~ Salary, data = merged_nba_clean)

# Extract residuals and predicted values from the model
residuals <- resid(model)
predicted_values <- fitted(model)

# Create a residual plot
plot(predicted_values, residuals,
     main = "Residual Plot of Individual Salaries",
     xlab = "Predicted Values",
     ylab = "Residuals")
abline(h = 0, col = "red")  # Add a horizontal line at y = 0 for reference

```
Resiudals seems very spread out and do not present an obvious pattern for the most part, so we can assume normality for a linear relationship between individual salaries and team win percentages.
```{r}
# Generate Q-Q plot to test for normality of 'Salary' data
qqnorm(merged_nba_clean$Salary, main = "Q-Q Plot of Individual Salaries")
qqline(merged_nba_clean$Salary, col = "red")
```
Q-Q plot on the other hand is not very good for assuming normality, we will take a look at other transformations below to see if any are any better, we will take a look at the correlation too. If the correlation is better then we will use that transformation.
```{r}
library(ggplot2)
# Perform square root transformation
merged_nba_clean$Salary_sqrt <- sqrt(merged_nba_clean$Salary)
# Fit a linear regression model using square root transformation
sqrt_model <- lm(WP ~ Salary_sqrt, data = merged_nba_clean)

# Print summary of the model
summary(sqrt_model)

# Plot the relationship between Total spending (sqrt) and WP with regression line
ggplot(merged_nba_clean, aes(x = Salary_sqrt, y = WP)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Square Root Transformation of Individual Salaries vs. Win Percentage",
       x = "Square Root of Salaries",
       y = "Win Percentage") +
  theme_minimal()

```
```{r}
residuals_sqrt <- residuals(sqrt_model)

# Create a residual plot
plot(merged_nba_clean$Salary_sqrt, residuals_sqrt,
     main = "Residual Plot of Square-Root Transformed Salaries",
     xlab = "Square Root of Individual Salaries",
     ylab = "Residuals")
abline(h = 0, col = "red")  # Add a horizontal line at y = 0 for reference
```

Residual plot does not look to bad, no pattern is apparent.
```{r}
# Create Q-Q plot of the residuals
qqnorm(residuals_sqrt, main = "Q-Q Plot of Residuals (Square Root Transformed Model)")
qqline(residuals_sqrt, col = "red")
```
Q-Q plot for square root transformation is definitely better than the orignal linear model, although still could be better.

Lets compare correlation between the original and square root transformation
```{r}
correlation1 <- cor(merged_nba_clean$Salary, merged_nba_clean$WP)


print(paste("Correlation between individual salary and team win percentage:", correlation1))
```
Since the square root transformation for invdividual salaries is higher than the default, we will use the square root transformation moving forward.
```{r}
correlation_sqrt <- cor(merged_nba_clean$Salary_sqrt, merged_nba_clean$WP)


print(paste("Correlation between individual salary(square-root transformation) and team win percentage:", correlation_sqrt))
```


First we will do correlation between individual player salaries and their teams' win percentage without clustering to see the default results.
```{r}
library(ggplot2)

correlation1 <- cor(merged_nba_clean$Salary_sqrt, merged_nba_clean$WP)


print(paste("Correlation between individual salary and team win percentage:", correlation1))

# Plot the relationship between individual salary and team win percentage
ggplot(merged_nba_clean, aes(x = Salary_sqrt, y = WP)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Correlation between Individual Salary and Team Win Percentage",
       x = "Individual Salary",
       y = "Win Percentage") +
  theme_minimal()
```

Now we will cluster by player salaries
```{r}
# load necessary libraries
library(dplyr)
library(ggplot2)
library(factoextra)

# Select the salary column
salary_data <- merged_nba_clean %>% select(Salary_sqrt)

# Normalize the salary data
salary_data <- scale(salary_data)

# Elbow method to find the optimal number of clusters(as researched)
fviz_nbclust(salary_data, kmeans, method = "wss") +
  geom_vline(xintercept = 3, linetype = 2) +
  labs(subtitle = "Elbow method")

# Set seed for reproducibility
set.seed(123)

# Apply K-means clustering
kmeans_result <- kmeans(salary_data, centers = 3, nstart = 25)

# Add cluster assignment to the original dataset
merged_nba_clean$cluster <- kmeans_result$cluster

# Plot the clusters
ggplot(merged_nba_clean, aes(x = Salary_sqrt, y = Salary_sqrt, color = factor(cluster))) +
  geom_point() +
  labs(title = "K-means Clustering of Players by Salary",
       x = "Salary_sqrt",
       y = "Salary_sqrt",
       color = "Cluster") +
  theme_minimal()

```
```{r}
# Count the number of players in each cluster
cluster_counts <- merged_nba_clean %>%
  group_by(cluster) %>%
  summarise(count = n())

print(cluster_counts)
```
Correlation between individual player salaries and their teams' win percentage with clustering to see more reliable results.
```{r}
# Calculate average salary per cluster
cluster_stat <- merged_nba_clean %>%
  group_by(cluster) %>%
  summarise(mean_salary = mean(Salary_sqrt), mean_WP = mean(WP))

print(cluster_stat)

# Calculate the correlation between mean spending and mean WP for clusters
correlation_cluster_salary_WP <- cor(cluster_stat$mean_salary, cluster_stat$mean_WP)

print(paste("Correlation between cluster mean salary and win percentage:", correlation_cluster_salary_WP))
```
```{r}
# Fit a linear regression model
model1 <- lm(WP ~ Total, data = merged_nba_clean)

# Extract residuals and predicted values from the model
residuals <- resid(model1)
predicted_values <- fitted(model1)

# Create a residual plot
plot(predicted_values, residuals,
     main = "Residual Plot of Team Spending",
     xlab = "Predicted Values",
     ylab = "Residuals")
abline(h = 0, col = "red")  # Add a horizontal line at y = 0 for reference
```
Residual plot looks pretty good for Team Spending, no apparent pattern, we can assume normality from this.

```{r}
# Generate Q-Q plot to test for normality of 'Total' data
qqnorm(merged_nba_clean$Total, main = "Q-Q Plot of Team Spending")
qqline(merged_nba_clean$Total, col = "red")
```
Once again, the Q-Q plot does not demonstrate a normal distribution, so we will look at different transformations.

```{r}
# Perform square root transformation
merged_nba_clean$Total_sqrt <- sqrt(merged_nba_clean$Total)
# Fit a linear regression model using square root transformation
sqrt_model1 <- lm(WP ~ Total_sqrt, data = merged_nba_clean)

# Print summary of the model
summary(sqrt_model1)

# Plot the relationship between Total spending (sqrt) and WP with regression line
ggplot(merged_nba_clean, aes(x = Total_sqrt, y = WP)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Sqaure Root Transformation of Team spending vs. Win Percentage",
       x = "Square Root Transformation of Team Spending",
       y = "Win Percentage") +
  theme_minimal()

```
```{r}
residuals_sqrt1 <- residuals(sqrt_model1)

# Create a residual plot
plot(merged_nba_clean$Total_sqrt, residuals_sqrt1,
     main = "Residual Plot of Square-Root Transformed Total Spending",
     xlab = "Square Root of Total Spending",
     ylab = "Residuals")
abline(h = 0, col = "red")  # Add a horizontal line at y = 0 for reference
```
Residuals have no apparent pattern so passes the check for normality. 

```{r}
# Create Q-Q plot of the residuals
qqnorm(residuals_sqrt1, main = "Q-Q Plot of Residuals (Exponential Model)")
qqline(residuals_sqrt1, col = "red")

```
The Q-Q plot is still not that good when using square root transformation on Team Spending. However we will compare it with the default to see which model we will proceed with.

```{r}
correlation2 <- cor(merged_nba_clean$Total, merged_nba_clean$WP)

print(paste("Correlation between Average Team Spending and Win Percentage Without Clustering:", correlation2))
```
```{r}
sqrt_corr2 <- cor(merged_nba_clean$Total_sqrt, merged_nba_clean$WP)

# Step 4: Print the correlation
print(paste("Correlation between Average Team Spending(Square Root Transformation) and Win Percentage Without Clustering:", sqrt_corr2))
```
Once again, the correlation for square root transformation was higher so we will use it to find the correlation between Team Spending and Win Percentage with and without clustering.

Correlation between average team spending and win percentage without clustering.
```{r}
library(ggplot2)


sqrt_corr2 <- cor(merged_nba_clean$Total_sqrt, merged_nba_clean$WP)


print(paste("Correlation between Average Team Spending(Square-Root Transformation) and Win Percentage Without Clustering:", sqrt_corr2))


ggplot(merged_nba_clean, aes(x = Total_sqrt, y = WP)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Correlation between Team Spending and Win Percentage",
       x = "Average Team Spending",
       y = "Win Percentage") +
  theme_minimal()

```
Cluster by Team Spending
```{r}
# load necessary libraries
library(dplyr)
library(ggplot2)
library(factoextra)

# Select the Total column
Total_data <- merged_nba_clean %>% select(Total_sqrt)

# Normalize the Total data
Total_data <- scale(Total_data)

# Elbow method to find the optimal number of clusters(as researched)
fviz_nbclust(Total_data, kmeans, method = "wss") +
  geom_vline(xintercept = 3, linetype = 2) +
  labs(subtitle = "Elbow method")

# Set seed for reproducibility
set.seed(23)

# Apply K-means clustering
kmeans_result1 <- kmeans(Total_data, centers = 3, nstart = 25)

# Add cluster assignment to the original dataset
merged_nba_clean$cluster1 <- kmeans_result1$cluster

# Plot the clusters
ggplot(merged_nba_clean, aes(x = Total, y = Total, color = factor(cluster1))) +
  geom_point() +
  labs(title = "K-means Clustering of Players by Team Spending",
       x = "Total",
       y = "Total",
       color = "Cluster") +
  theme_minimal()

```



We will now find correlation between average team spending and win percentage with clustering.
```{r}
# Calculate the mean salary and WP for each cluster
cluster_stats <- merged_nba_clean %>%
  group_by(cluster1) %>%
  summarise(mean_spending = mean(Total_sqrt), mean_WP = mean(WP))

print(cluster_stats)

# Calculate the correlation between mean spending and mean WP for clusters
correlation_cluster_spending_WP <- cor(cluster_stats$mean_spending, cluster_stats$mean_WP)

print(paste("Correlation between cluster mean salary and win percentage:", correlation_cluster_spending_WP))
```
We look at individual points since total points are what decide the outcome of a game. So we correlate a player's average points with their salary.
```{r}
library(ggplot2)

correlation_pts <- cor(merged_nba_clean$PTS, merged_nba_clean$Salary)

print(paste("Correlation between individual Points and Salary:", correlation_pts))

ggplot(merged_nba_clean, aes(x = PTS, y = Salary)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Correlation between Average Points and Salary",
       x = "Average Points",
       y = "Salary") +
  theme_minimal()

```
```{r}
library(dplyr)
library(ggplot2)
library(factoextra)

# Select the Points column
points_data <- merged_nba_clean %>% select(PTS)

# Normalize the Points data
points_data <- scale(points_data)

# Elbow method to find the optimal number of clusters
fviz_nbclust(points_data, kmeans, method = "wss") +
  geom_vline(xintercept = 3, linetype = 2) +
  labs(subtitle = "Elbow method")

# Set seed for reproducibility
set.seed(123)

# Apply K-means clustering
kmeans_result_points <- kmeans(points_data, centers = 3, nstart = 25)

# Add cluster assignment to the original dataset
merged_nba_clean$cluster_points <- kmeans_result_points$cluster

# Plot the clusters
ggplot(merged_nba_clean, aes(x = PTS, y = WP, color = factor(cluster_points))) +
  geom_point() +
  labs(title = "K-means Clustering of Players by Points",
       x = "Points",
       y = "Win Percentage",
       color = "Cluster") +
  theme_minimal()

# Count the number of players in each cluster
cluster_counts_points <- merged_nba_clean %>%
  group_by(cluster_points) %>%
  summarise(count = n())

print(cluster_counts_points)
```
```{r}
# Calculate the mean salary and WP for each cluster
cluster_stats1 <- merged_nba_clean %>%
  group_by(cluster_points) %>%
  summarise(mean_points = mean(PTS), mean_salaries = mean(Salary))

print(cluster_stats1)

# Calculate the correlation between mean spending and mean WP for clusters
correlation_cluster_PTS_Salary <- cor(cluster_stats1$mean_points, cluster_stats1$mean_salaries)

print(paste("Correlation between cluster mean points and Salary:", correlation_cluster_PTS_Salary))
```
Conclusion: Based on the several questions that we generated, we have concluded that NBA Players' salaries are justified. With high correlation of 0.90 + on all 3 questions and clustering we were able to gather that NBA players with similar salaries are paid well based on their individual performances and their effect on team win percentages.




